Versioned data: why it is needed and how it can be achieved (easily and cheaply)

### 2017-09-22 ms.tex
Fix small bug preventing compile
wording changes throughout

As is evidenced in the advent of this journal, publication of quality datasets is now considered a first-class scientific product. Increasingly, funding bodies, publishers, and scientific social norms are recognizing the value of sharing data, including as standalone products without any accompanying analyses \cite[e.g.][]{Whitlock-2011,Fairbairn-2011,Piwowar-2011,VanNoorden-2013,Gibney-2013,Editorial-2014}. Datasets are now routinely archived as part of the publishing process. Moreover, increasing numbers of standalone ``Data papers'' (or descriptors) have been appearing in standard domain-level journals, as well as specialised data journals, like \emph{Scientific Data}. Yet, while the last decade has witnessed a rapid and exciting change in attitudes towards data sharing and publishing, the scientific community is still grappling with how to effectively disseminate and manage open-source datasets \cite{Whitlock-2011, Goodman-2014, Lowndes-2017,Perkel-2016,VanNoorden-2013, Kratz-2015}. In particular, the current publishing model has not yet embraced the idea that interesting datasets are constantly evolving, i.e. never "finished".

Most of the datasets published in this journal -- or other similar data descriptors -- are likely to be living entities, meaning the canonical (or master) version of the dataset changes through time, usually for the better. Typical changes involve: the addition of new data, improving the quality of existing data, or structuring of the database content. For example a dataset on a biological organisms might be expanded through the addition of new records or improved through the correction of spelling mistakes in taxonomic names. As research around a data product grows, there might be many such additions. Ideally, any changes to a dataset would become immediately available to any interested parties, while -- to ensure reproducibility -- past versions of the dataset would remain accessible.

Unfortunately, current models for publishing datasets do not immediately solve the issue of of how to distribute successive versions of a dataset in an effective and scalable manner. The most common current working model for publishing scientific datasets is to archive them in a variety of open-source data repositories -- such as \texttt{Dryad}, \texttt{Figshare}, or \texttt{Zenodo} -- where the data is attributed a Digital Object Identifier (\textsc{doi}). Once published, the datasets are immutable. While future updates may be published under an updated a separate \textsc{doi}, the \textsc{doi} system does not readily connect different versions of a dataset. That means that most published datasets exist as a singular snapshot of the dataset, taken at the itme the data descriptor was published. While it might only require a small amount of work to update the dataset, if there is no way to distribute the updated dataset, then that work has to be re-done by every user of that dataset. Moreover, those required changes add steps between the canonical dataset and any analysis using it, making reproducibility of the new analysis more difficult than it needs to be.

Many research groups solve the issue of updating data in a variety of suboptimal ways. A common solution is to email around the latest versions. Another approach is to repeat corrections or updates that have been made elsewhere. This practice is both inefficient and a barrier to reproducibility, as the dataset used in a publication may differ to the version previously made available on-line. There is therefore a strong need for an easy way to distribute changes to a dataset after it's initial publication to potential users, along with notes on what has changed and why since the previous version.

In this article we i) introduce the concept of Versioned Data Delivery (\textsc{vdd}); ii) outline how emerging technologies in data science (Table \ref{tab:technologies}) can be used to help researchers maintain, distribute, and access small-to-medium sized datasets; and iii) introduce a new \textsc{R} package called \texttt{datastorr}, a proof-of-concept implementation of a \textsc{vdd} system. The issue of updating and expanding published data has been partly addressed in large centralised repositories like genetic sequences (\texttt{GenBank}) or species location data (\texttt{GBIF}), where new data can be added and there exist abilities to correct errors in existing records (e.g. by adding multiple version of a genetic sequence). Yet, such large centralised repositories require a level of infrastructure that is beyond most research groups, or even countries (Table \ref{tab:sql_v_versioneddata}). Our focus here is on the wide range of datasets not covered by these repositories, such as those appearing in this journal. In most of these cases, the data
collected will not be ``big'' but rather small-to-medium sized. Such data
support research projects, but are not in most cases general enough to
establish bespoke infrastructure for. Further, While we emphasize particular technologies in our implementation, the principles are general and could easily be ported to other platforms. Likewise, while the examples cited here come from our own research areas, ecology and evolutionary biology, the issues we have identified and the solutions we have put forward are relevant to nearly all data-oriented scientific domains.

@@

% So, git doesn't actually work this way, though everyone seems to
% think that it does.  Git stores *content* and then later tries to
% de-duplicate it.  It then displays changes as line-by-line diffs
% (but you can change this to display the changes other ways; see
% --color-words --ignore-all-space for example.  The problem is that
% the algorithms that git uses are optimised for typical sizes of code
% and not for data.  Small data is actually totally fine to store with
% git.  The other issue that turns up is if the files are _binary_
% then an arbitrarily small change in the underlying data can result
% in an arbitrarily large change in the binary representation, which
% prevents deduplication.
%
% Git-LFS tries to step around this by storing large files separately
% from git's general machinery.  Git-annex did the same.  But they
% suffer from some of the other issues with git which are that it's a
% bit of a pain for (especially novice) users to do nontrivial things
% with.  I _always_ have to look up how to compare two versions across
% branches.  And there's no simple way of loading in two versions of
% the data into R without ending up with two working copies or doing
% some nasty "git show" call.
Because \texttt{git} was built for tracking changes in code, there are aspects of how it behaves that are not natural to apply to  data \cite{Perkel-2016}.  For example, because it tracks changes line-by-line rather than cell-by-cell, there are inefficiencies in how the software stores the changes.
% Dan Noble: I think this is a pragmatic view, but I would have a tendency not to discuss how the tools you can use for a VDD approach are likely to be "replaced in the near future”. Of course, things change, but this will not entice the people you want the paper to appeal to to adopt this sort of approach. From my experience, some of the resistance to adopting this stuff comes from programs or repositories being “replaced” at such a high rate. They just don’t feel like its worth doing.
There is rapid current work on this problem and prototypes available, and as such, it is likely that \texttt{git} for data will replaced by a more natural system for data in the near future  \citep{Fli, Dat}.  This future system would be faster computationally, smaller in file size, flexible with respect to data structures, and thus allow for larger datasets to be versioned efficiently.  That said, the workflow we describe here currently works well with \texttt{git} and should work even more quickly with its future replacement.

@@

@@

@@

@@

The problem of tracking data through time and versions is likely to be widespread. There are many alternate solutions, with associated features and drawbacks.
% Reword
The current structure of scientific research is very distributed, with research groups across the world working on the same question.  Moreover at a time of funding scarcity, many world-class research groups do not have excess funds to hire and support technology specialists.  As such we argue that there is a niche for a easy, cheap, and flexible \textsc{vdd} system.

@@

\begin{table}[h!]
\centering
\caption{Groups of users interacting with the Versioned Data Delivery system described and their requirements}
{\footnotesize
\vspace{1cm}
  \begin{tabular}{p{2cm}p{5cm}p{7cm}}
  \hline
% Latex being the worst has made the overflow in the second column flow through to affect the third.  I used to know how to deal with this...
  \textbf{Group} & \textbf{Goal} & \textbf{Requirements} \\ \hline
  Maintainer & Create and distribute versioned datasets & Low technical overhead \\
    & & Easy workflow for releasing new versions \\
    & & Long term preservation \\
    & & Easy to crowd-source error checking and contributions \\
    & & Low initial cost \\
    & & Low on-going maintenance \\
  Contributor & Contribute to future versions of a dataset & Add new data \\
    & & Report errors in existing data  \\
  Users (all) & Easy access to releases from a versioned dataset & Introduction \& overview \\
    & & Long term stability \\
    & & Clear path for users to become contributors \\
  Users (code-based) & Build reproducible products using specific versions of a dataset & Scripted access to releases\\
    & & Easy installation\\
    & & Long term stability \\
  \hline
  \end{tabular}
  % }
\label{tab:user_requirements}
\end{table}

>>>>

As is evidenced in the advent of this journal, publication of quality datasets is now considered a first-class scientific product. Increasingly, funding bodies, publishers, and scientific social norms are recognizing the value of sharing data, including as standalone products without any accompanying analyses \cite[e.g.][]{Whitlock-2011,Fairbairn-2011,Piwowar-2011,VanNoorden-2013,Gibney-2013,Editorial-2014}. Datasets are now routinely archived as part of the publishing process. Moreover, increasing numbers of standalone ``Data papers'' (or descriptors) have been appearing in standard domain-level journals, as well as specialised data journals, like \emph{Scientific Data}. Yet, while the last decade has witnessed a rapid and exciting change in attitudes towards data sharing and publishing, the scientific community is still grappling with how to effectively disseminate and manage open-source datasets \cite{Whitlock-2011, Goodman-2014, Lowndes-2017,Perkel-2016,VanNoorden-2013, Kratz-2015}. In particular, the current data publishing model has not yet embraced the idea that many  datasets are designed to answer scientific questions that extend beyond the scope of a single empirical paper.  These datasets are constantly evolving, i.e. never "finished", and may be the subject of repeated analysis or meta-analysis.

Many, and perhaps most, of the datasets published in this journal are likely to be living entities, meaning the state-of-the-art version of the dataset will changes through time. Typical changes may involve: the addition of new data, improving the quality of existing data, better linked to external databases, or re-structuring of the database content to be able to address new questions. For example a dataset on a biological organisms might be expanded through the addition of new records or improved through the correction of spelling mistakes in taxonomic names. As research around a data product grows, there might be many such additions. Ideally, any changes to a dataset would become immediately available to any interested parties, while -- to ensure reproducibility -- past versions of the dataset would remain accessible.

Unfortunately, current models for publishing datasets do not immediately solve the issue of how to distribute successive versions of a dataset in an effective and scalable manner. The most common current working model for publishing scientific datasets is to archive them in a variety of open-source data repositories -- such as \texttt{Dryad}, \texttt{Figshare}, or \texttt{Zenodo} -- where the data is attributed a Digital Object Identifier (\textsc{doi}). Once published, the datasets are immutable. That means that most published datasets exist as a singular snapshot of the dataset, taken at the time the data descriptor was published. While it might only require a small amount of work to update the dataset, if there is no way to distribute the updated dataset, then that work has to be re-done by every user of that dataset. Moreover, those required changes add steps between the canonical dataset and any analysis using it, making reproducibility of the new analysis more difficult than it needs to be.

Many research groups internally solve the issue of versioning data in a variety of suboptimal ways. A common solution is to email around the latest versions with version numbers appended to the filename. Another approach is to repeat corrections or updates that have been made elsewhere. This practice is both inefficient and a barrier to reproducibility, as the dataset used in a publication may differ to the version previously made available on-line. There is therefore a strong need for an easy way to distribute changes to a dataset after its initial publication to potential users, along with notes on what has changed and why since the previous version.

In this article we i) introduce the concept of Versioned Data Delivery (\textsc{vdd}); ii) outline how emerging technologies in data science (Table \ref{tab:technologies}) can be used to help researchers maintain, distribute, and access small-to-medium sized datasets; and iii) introduce a new \textsc{R} package called \texttt{datastorr}, a proof-of-concept implementation of a \textsc{vdd} system. The issue of updating and expanding published data has been partly addressed in large centralised repositories like genetic sequences (\texttt{GenBank}) or species location data (\texttt{GBIF}), where new data can be added and there exist abilities to correct errors in existing records (e.g. by adding multiple version of a genetic sequence). Yet, such large centralised repositories require a level of infrastructure that is beyond most research groups, or even countries (Table \ref{tab:sql_v_versioneddata}). Our focus here is on the wide range of datasets not covered by these repositories, such as those appearing in this journal. In most of these cases, the data collected will not be ``big'' but rather small-to-medium sized. Such data support important research projects on particular scientific questions, but are not in most cases general enough to warrant custom infrastructure. Further, while we emphasize particular technologies in our implementation, the principles are general and could easily be ported to other platforms. Likewise, while the examples cited here come from our own research areas the issues we have identified and the solutions we have put forward are relevant to nearly all data-oriented scientific domains.

@@

% So, git doesn't actually work this way, though everyone seems to
% think that it does.  Git stores *content* and then later tries to
% de-duplicate it.  It then displays changes as line-by-line diffs
% (but you can change this to display the changes other ways; see
% --color-words --ignore-all-space for example.  The problem is that
% the algorithms that git uses are optimised for typical sizes of code
% and not for data.  Small data is actually totally fine to store with
% git.  The other issue that turns up is if the files are _binary_
% then an arbitrarily small change in the underlying data can result
% in an arbitrarily large change in the binary representation, which
% prevents deduplication.
%
% Git-LFS tries to step around this by storing large files separately
% from git's general machinery.  Git-annex did the same.  But they
% suffer from some of the other issues with git which are that it's a
% bit of a pain for (especially novice) users to do nontrivial things
% with.  I _always_ have to look up how to compare two versions across
% branches.  And there's no simple way of loading in two versions of
% the data into R without ending up with two working copies or doing
% some nasty "git show" call.
Because \texttt{git} was built for tracking changes in code, there are aspects of how it behaves that are not natural to apply to  data \cite{Perkel-2016}.  For example, because it tracks changes line-by-line rather than cell-by-cell, there are inefficiencies in how the software stores the changes.
% Dan Noble: I think this is a pragmatic view, but I would have a tendency not to discuss how the tools you can use for a VDD approach are likely to be "replaced in the near future”. Of course, things change, but this will not entice the people you want the paper to appeal to to adopt this sort of approach. From my experience, some of the resistance to adopting this stuff comes from programs or repositories being “replaced” at such a high rate. They just don’t feel like its worth doing.
There is rapid current work on new implementations which will be faster computationally, smaller in file size, flexible with respect to data structures, and thus allow for larger datasets to be versioned efficiently\citep{Fli, Dat}.
% Add CKAN here?
That said, the workflow we describe here currently works well with \texttt{git} and should work even more quickly with its future replacement.

@@

% WKC: I agree. Not sure what to do with these two sections which are a bit of a tangent relative to our main point

@@

% WKC: this is back to our main point

@@

% WKC: hmmmm not sure we've nailed this yet, but see
% https://github.com/traitecoevo/data_versioning/issues/10#issuecomment-317059819

@@

%work on this paragraph more
The problem of tracking data through time and versions is likely to be widespread. There are many alternate solutions, with associated features and drawbacks.  Some large data distribution institutions have begon grabling with this issue in various ways (see \url{https://www.ncbi.nlm.nih.gov/genbank/sequenceids/}), but the current structure of scientific research is very distributed, with research groups across the world working on the same question.  Moreover at a time of funding scarcity, many world-class research groups do not have excess funds to hire and support technology specialists.  As such we argue that there is a niche for a easy, cheap, and flexible \textsc{vdd} system.

@@

\begin{table}[h!]
\centering
\caption{Groups of users interacting with the Versioned Data Delivery system described and their requirements}
{\footnotesize
\vspace{1cm}
  \begin{tabular}{p{2cm}p{5cm}p{7cm}}
  \hline
% Latex being the worst has made the overflow in the second column flow through to affect the third.  I used to know how to deal with this...
  \textbf{Group} & \textbf{Goal} & \textbf{Requirements} \\ \hline
  Maintainer & Create and distribute versioned datasets & Low technical overhead \\
    & & Easy workflow for releasing new versions \\
    & & Long term preservation \\
    & & Easy to crowd-source error checking and contributions \\
    & & Low initial cost \\
    & & Low on-going maintenance \\
  Contributor & Contribute to future versions of a dataset & Add new data \\
    & & Report errors in existing data  \\
  Users (all) & Easy access to releases from a versioned dataset & Introduction \& overview \\
    & & Long term stability \\
    & & Clear path for users to become contributors \\
  Users (code-based) & Build reproducible products using specific versions of a dataset & Scripted access to releases\\
    & & Easy installation\\
    & & Long term stability \\
  \hline
  \end{tabular}
}
\label{tab:user_requirements}
\end{table}
